<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# *methylTFR* : Quantification of DNA Methylation Patterns in TFBS

<!-- badges: start -->
[![Test R-universe](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/r-universe-test.yml/badge.svg)](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/r-universe.yml)
[![R-CMD-check-bioc](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/check-bioc.yml/badge.svg)](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/check-bioc.yml)
[![R-CMD-check](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/R-CMD-check.yaml)
[![GitHub issues](https://img.shields.io/github/issues/EpigenomeInformatics/methylTFR)](https://github.com/EpigenomeInformatics/methylTFR/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/EpigenomeInformatics/methylTFR)](https://github.com/EpigenomeInformatics/methylTFR/pulls)
<!-- badges: end -->

`methylTFR` is an R-package to analyze DNA methylation signatures in transcription factor binding sites in each individual cells or samples.

## Installation instructions

Get the latest release `methylTFR` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("methylTFR")
```

And the development version from [GitHub](https://github.com/EpigenomeInformatics/methylTFR) with:

```{r 'install_dev', eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_github("EpigenomeInformatics/methylTFR")
```
## Documentation

Full documentation and vignettes are hosted at
[epigenomeinformatics.github.io/methylTFR](https://epigenomeinformatics.github.io/methylTFR/):

- [Get started](https://epigenomeinformatics.github.io/methylTFR/articles/methylTFR.html) — reading data, computing deviations, and footprints.
- [Case study: memory vs. naive T cells](https://epigenomeinformatics.github.io/methylTFR/articles/memTcells.html) — differential TF activity on bundled example data.

## Quick Start

This is a basic example which shows you how to run `methylTFR` :

```{r example, eval = FALSE}
library(GenomicRanges)
library(dplyr)
library(methylTFRAnnotationHg38) # annotation package for hg38
library(methylTFR)

gcfreqs <- getGCfreq(motifSet = "jaspar2020")
gc_dist <- getGenomeGC("hg38")
tf_bindsites <- getTFbindsites(motifSet = "jaspar2020")

sample_dir <- file.path("samples_dir")
sample_ann <- "samples.tsv" # should contain column name bedFile

# deviation score matrix
deviations <- run_methyltfr(sample_ann, # sample annotation file
    sample_dir, # where the EPP files are
    threads = 8, # number of threads
    chunkSize = 10, # number of chunks to process
    sampleColName = "bedFile", # column name for EPP file paths in sample_ann
    tf_bindsites = tf_bindsites, # TF binding sites
    gcfreqs = gcfreqs, # GC frequency
    gc_dist = gc_dist, # GC distribution
    filetype = "EPP" # file type
)
```

## Citation

Below is the citation output from using `citation('methylTFR')` in R. Please
run this yourself to check for any updates on how to cite __methylTFR__.

```{r 'citation', eval = requireNamespace('methylTFR')}
print(citation("methylTFR"), bibtex = TRUE)
```
