---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# methylTFR

<!-- badges: start -->
[![R-CMD-check-bioc](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/check-bioc.yml/badge.svg)](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/check-bioc.yml)
[![R-CMD-check](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EpigenomeInformatics/methylTFR/actions/workflows/R-CMD-check.yaml)
[![GitHub issues](https://img.shields.io/github/issues/EpigenomeInformatics/methylTFR)](https://github.com/EpigenomeInformatics/methylTFR/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/EpigenomeInformatics/methylTFR)](https://github.com/EpigenomeInformatics/methylTFR/pulls)
<!-- badges: end -->

# methylTFR

`methylTFR` is an R-package to analyze DNA methylation patterns on transcription factor binding sites in each individual cells or samples.
It takes inputs from `RnBeads` processed bed file, *EPP* format. 

## Installation instructions

Get the latest release `methylTFR` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("methylTFR")
```

And the development version from [GitHub](https://github.com/EpigenomeInformatics/methylTFR) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("EpigenomeInformatics/methylTFR")
```
## Example

This is a basic example which shows you how to run `methylTFR` :

```{r example, eval = FALSE}
library(GenomicRanges)
library(dplyr)
library(methylTFRAnnotationHg38)
library(methylTFR)

gcfreqs <-  getGCfreq()
gc_dist <- getGenomeGC()
tf_bindsites <- getTFbindsites()

sample_dir <- file.path("samples_dir")
sample_ann <- "samples.tsv" # should contain column name bedFile

# deviation score matrix
deviations <- run_methyltfr(sample_ann,
                            sample_dir,
                            threads = 8,
                            threads = 30,
                            chunkSize = 10,
                            sampleColName = "bedFile",
                            genome = "hg38",
                            tf_bindsites = tf_bindsites,
                            gcfreqs = gcfreqs,
                            gc_dist = gc_dist,
                            filetype = "EPP")

```

## Citation

Below is the citation output from using `citation('methylTFR')` in R. Please
run this yourself to check for any updates on how to cite __methylTFR__.

```{r 'citation', eval = requireNamespace('methylTFR')}
print(citation('methylTFR'), bibtex = TRUE)
```

