---
title: "Build methylTFR annotations"
author: 
  - name: Irem B. Gunduz
    affiliation:
    - Saarland University
    email: irembgunduz@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('methylTFR')`"
vignette: >
  %\VignetteIndexEntry{build methylTFR annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
  R = citation(),
  BiocStyle = citation("BiocStyle")[1],
  knitr = citation("knitr")[1],
  RefManageR = citation("RefManageR")[1],
  rmarkdown = citation("rmarkdown")[1],
  sessioninfo = citation("sessioninfo")[1],
  testthat = citation("testthat")[1],
  methylTFR = citation("methylTFR")[1]
)
```

## Install `methylTFR`

`r Biocpkg("methylTFR")` is available from Bioconductor. To install it, please follow the instructions in the [Bioconductor landing page](https://bioconductor.org/packages/methylTFR/). In particular, you will need to install the latest version of `R` and `BiocManager` and then run the following commands:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("methylTFR")
```

## What is `methylTFR` annotations?

`methylTFR` computes expected methylation values using GC content of the genome. Annotation builder functions are part of the `methylTFR` package.
In nuthsell, you will need a two component to build your own annotation:

- A genome object, such as `BSgenome.Hsapiens.UCSC.hg19` from the `BSgenome` package.
- A desired motif annotations, such as `JASPAR2020`

### Quick start to build your own annotations using `methylTFR`
Below we illustrate how to build your own annotation using `BSgenome.Hsapiens.UCSC.hg38` as a genome for `JASPAR2020` motifs. 

```{r "quickstart", message=FALSE, eval=FALSE}
suppressPackageStartupMessages({
  library("BSgenome.Hsapiens.UCSC.hg38")
  library("methylTFR")
  library("JASPAR2020")
  library("TFBSTools")
  library("logger")
})

# compute genome wise GC bias and save it as a GRanges object
genomewideGC <- computeGenomeWideGC(BSgenome.Hsapiens.UCSC.hg38,
  onlyMain = TRUE, includeSexChr = TRUE, num_cores = 1
)


# Get Jaspar2020 motif PFM matrix list
motifPFMatrixList <- getMatrixSet(
  x = JASPAR2020,
  opts = list(species = 9606, all_versions = FALSE)
)

# TODO add a function to convert motif PFM matrix list to GRanges object

log_start("Calculating GC distribution ...")
gc_dist <- calculate_gcdist(genome = BSgenome.Hsapiens.UCSC.hg38, threads = 1)
log_info("Applying quantile ...")
gc_bin <- quantile(gc_dist, probs = seq(0, 1, 1 / 5))
log_success("GC distribution calculated.")

motif_list <- names(tf_bindsites)
logger.info(paste0("Number of motifs within the annotation: ", length(motif_list)))

logger_start("Generating GC matrix for each motif ...")
motif_gcfreq <- mclapply(chunk_motifs, processMotifs2GCMatrix,
  gc_bin = gc_bin, genome = BSgenome.Hsapiens.UCSC.hg38,
  mc.cores = 1
)
logger_success("Motifs processed.")
```
After you succesfully build your annotation, you can use it with `methylTFR` functions.

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `methylTFR` tag and check [the older posts](https://support.bioconductor.org/tag/methylTFR/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

## Citing `methylTFR`

We hope that `r Biocpkg("methylTFR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("methylTFR")
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```
