---
title: "Case study: TF activity in memory vs. naive T cells"
author: "Irem B. Gündüz, Sarath Kumar Murugan, Fabian Muller"
date: "`r Sys.Date()`"
package: methylTFR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Case study: TF activity in memory vs. naive T cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Overview

CD4+ T cells reshape their regulatory programs as they differentiate from
a naive to a memory state. This case study uses `methylTFR` deviation
scores to ask a focused question: **which transcription factors show
differential activity between memory and naive T cells?**

We reuse deviation matrices bundled with the package (`tc_mem`, `tc_naive`),
so the analysis runs end-to-end without downloading the *hg38* annotation.
See the [Get started](methylTFR.html) vignette for how deviations are
computed from raw WGBS data.

```{r init}
library(methylTFR)
library(ggplot2)
```

## Deviation data

Each object is a `methylTFRdeviations` matrix of motifs (rows) by
samples (columns). We combine the two groups into a single matrix.

```{r load}
load(system.file("extdata", "tc_mem.rda", package = "methylTFR"))
load(system.file("extdata", "tc_naive.rda", package = "methylTFR"))

devs <- cbind(tc_mem, tc_naive)
devs
```

Group labels are taken from the sample names (the prefix before `_`).

```{r groups}
get_groupname <- function(x) unlist(strsplit(x, split = "_"))[1]
groups <- sub(
    ".bedGraph", "",
    vapply(colnames(devs), get_groupname, character(1))
)
table(groups)
```

## Differential TF activity

`differential_deviation_test()` compares deviation scores between the two
groups per motif. With two groups and `parametric = TRUE` this is a
t-test, followed by Benjamini-Hochberg correction.

```{r difftest}
tc_result <- differential_deviation_test(
    deviations = devs,
    groups = groups,
    alternative = "two.sided",
    parametric = TRUE,
    padjMethod = "BH"
)

tc_result <- tc_result[order(tc_result$p_value_adjusted), ]
tc_result
```

The `mean_difference` column is the absolute difference in mean deviation
between memory and naive cells: larger values flag TFs whose binding-site
methylation shifts most with differentiation.

## Ranking the top TFs

We visualise motifs by effect size, highlighting those passing an FDR
cutoff of 0.05.

```{r rankplot, fig.width = 6, fig.height = 4}
tc_result$motifs <- factor(
    tc_result$motifs,
    levels = tc_result$motifs[order(tc_result$mean_difference)]
)
tc_result$significant <- ifelse(
    tc_result$p_value_adjusted < 0.05,
    "FDR < 0.05", "n.s."
)

ggplot(tc_result, aes(x = mean_difference, y = motifs, colour = significant)) +
    ggplot2::geom_segment(
        aes(x = 0, xend = mean_difference, yend = motifs),
        colour = "grey70"
    ) +
    ggplot2::geom_point(size = 3) +
    ggplot2::scale_colour_manual(
        values = c("FDR < 0.05" = "#E66100", "n.s." = "grey50")
    ) +
    ggplot2::labs(
        x = "Mean deviation difference (memory vs. naive)",
        y = NULL, colour = NULL,
        title = "Differential TF activity"
    ) +
    theme_classic() +
    theme(legend.position = "bottom")
```

## Interpretation

TFs at the top of the ranking are the strongest candidates for driving
memory-specific regulatory changes: their binding sites lose or gain
methylation as cells transition out of the naive state. These hits are a
natural starting point for footprint inspection
(`plotExpectedFootprint()`, shown in the main vignette) or for follow-up
in a larger cohort.

Because the bundled data is a small subset (10 motifs, 5 samples per
group), the values here are illustrative. On the full BLUEPRINT memory
T-cell panel the same workflow scales to thousands of motifs via
`run_methyltfr()`.

## Session Information

```{r session-info}
sessionInfo()
```
