---
title: "Case study: TF activity in memory vs. naive T cells"
author: "Irem B. Gündüz, Sarath Kumar Murugan, Fabian Muller"
date: "`r Sys.Date()`"
package: methylTFR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Case study: TF activity in memory vs. naive T cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

This vignette demonstrates a complete downstream analysis of
`r Biocpkg("methylTFR")` deviation scores, starting from a precomputed
`methylTFRdeviations` object. It covers dimensionality reduction,
differential activity testing between two groups, comparison of two
independent contrasts, and inspection of the motif footprints that the
deviation scores summarise.

The biological setting is the naive-to-memory transition in human T
cells, in the cytotoxic (CD8, `Tc`) and helper (CD4, `Th`) compartments.
All quantities are derived from DNA methylation.

The following `r Biocpkg("methylTFR")` functionality is demonstrated:

| Step | Function |
|------|----------|
| Access bias-corrected deviation scores | `deviations()` |
| Access row-wise deviation Z-scores | `deviationZScores()` |
| Rank motifs by activity variability | `computeZScoreVariability()` |
| Test motifs between two groups | `differential_deviation_test()` |
| Plot observed and expected methylation | `plotExpectedFootprint()` |
| Plot the bias-corrected footprint | `plotMotifFootprint()` |

Deviation scores themselves are computed with `run_methyltfr()` from
per-sample methylation files, or with `run_methylTFR_RnBeads()` directly
from a preprocessed `RnBeads` object. Both are covered in the
[Get started](methylTFR.html) vignette; this one begins from their
output.

```{r init, message = FALSE}
library(methylTFR)
library(SummarizedExperiment)
library(ggplot2)
```

A single palette is used for the seven cell types throughout, so that a
colour denotes the same cell type in every figure.

```{r palette}
cell_type_colors <- c(
  "Th-Mem"   = "#41B6C4",
  "Tc-Mem"   = "#4292C6",
  "Tc-Naive" = "#888FB5",
  "Th-Naive" = "#C7E9B4",
  "B-cell"   = "#AE017E",
  "NK-cell"  = "#A65628",
  "Monocyte" = "#CC4C02"
)
```

# The example dataset

`immuneDeviations` contains bias-corrected deviation scores for
pseudobulk methylomes of seven human immune cell types, with one column
per donor sample and one row per JASPAR2020 motif. Four of the seven are
T cell subsets, naive and memory in the cytotoxic and helper
compartments; the remaining three provide a lineage contrast.

The underlying methylomes are from Gündüz *et al.* (2025); see the
References section.

```{r load}
load(system.file("extdata", "immuneDeviations.rda", package = "methylTFR"))
immuneDeviations

table(colData(immuneDeviations)$cell_type)
```

Provenance is stored in the object metadata.

```{r provenance}
metadata(immuneDeviations)
```

The source dataset provides 38 donor samples per cell type. To limit the
package size it was reduced on the sample axis only, to 15 samples per
cell type; every motif is retained. Subsetting motifs would be the more
obvious reduction, but selecting motifs by variability and then
demonstrating `computeZScoreVariability()` on the result would be
circular. The script used to generate the object is
`inst/scripts/tcell_data.R`.

## Accessing the two assays

A `methylTFRdeviations` object stores two matrices. `deviations()`
returns the bias-corrected deviation scores. `deviationZScores()`
returns the same values standardised row-wise across samples.

```{r assays}
dev_mat <- deviations(immuneDeviations)
z_mat <- deviationZScores(immuneDeviations)

round(dev_mat[seq_len(3), seq_len(4)], 4)
```

The two assays serve different purposes. Deviation scores are the
quantitative estimate of motif activity and are the correct input to
statistical tests. Row-wise Z-scores rescale every motif to a common
spread, which makes rows comparable in a heatmap, but by construction
the standard deviation of each row is 1.

```{r zscale}
summary(as.vector(dev_mat))
summary(apply(z_mat, 1, sd))
```

# 1. Dimensionality reduction

Deviation scores form a low-dimensional feature set, so principal
component analysis can be applied to the matrix directly.

```{r pca, fig.height = 4.2}
pca <- prcomp(t(dev_mat), center = TRUE, scale. = FALSE)
var_expl <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

pca_df <- data.frame(
  PC1 = pca$x[, 1], PC2 = pca$x[, 2],
  cell_type = colData(immuneDeviations)$cell_type
)

ggplot(pca_df, aes(PC1, PC2, colour = cell_type)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_manual(values = cell_type_colors, name = NULL) +
  labs(
    x = sprintf("PC1 (%.1f%%)", var_expl[1]),
    y = sprintf("PC2 (%.1f%%)", var_expl[2])
  ) +
  theme_classic()
```

PC1 separates the myeloid samples from the lymphoid ones. PC2 resolves
the T cell subsets by differentiation state: the two naive subsets group
together, as do the two memory subsets, while the CD8 and CD4
compartments overlap. Differentiation state therefore accounts for more
of the variation in motif activity than the CD8/CD4 distinction does.

# 2. Differential TF activity

## Ranking motifs by variability

Before testing a specific contrast, `computeZScoreVariability()` ranks
motifs by how much their activity varies across the dataset, without
requiring group labels. Each sample is calibrated against a null
estimated across motifs, so a variability above 1 indicates a motif
varying more than the background spread of that sample. P-values come
from a chi-squared test against that null.

```{r variability}
variability <- computeZScoreVariability(immuneDeviations, method = "robust")
variability <- variability[order(-variability$variability), ]

head(variability, 8)
```

The highest-ranking motifs belong to the CEBP family, which distinguish
the myeloid lineage from the lymphoid ones. This is a screen across all
cell types and requires no group labels, which makes it complementary to
the two-group test used below.

This function reads the `deviations` assay. Supplying row-wise Z-scores
instead would be uninformative, because their per-row standard deviation
is 1 for every motif, as shown above.

Optional bootstrap confidence bounds are available via
`bootstrap = TRUE`.

## Testing between two groups

`differential_deviation_test()` tests each motif for a difference in
deviation scores between two groups. With two groups and
`parametric = TRUE` the test is a Welch t-test; p-values are adjusted
with the Benjamini-Hochberg procedure.

The returned `mean_difference` column is unsigned, so the direction of
each change is computed separately. Differences below are oriented as
naive minus memory, so that a positive value indicates a higher
deviation score in the naive state.

```{r helper}
compare_subsets <- function(object, naive, memory) {
  grp_all <- colData(object)$cell_type
  keep <- grp_all %in% c(naive, memory)
  grp <- factor(as.character(grp_all[keep]), levels = c(naive, memory))

  dev_sub <- deviations(object)[, keep, drop = FALSE]
  z_sub <- deviationZScores(object)[, keep, drop = FALSE]
  is_naive <- grp == naive

  res <- differential_deviation_test(
    deviations = dev_sub,
    groups = grp,
    alternative = "two.sided",
    parametric = TRUE,
    padjMethod = "BH"
  )
  res$diff <- rowMeans(dev_sub[, is_naive, drop = FALSE]) -
    rowMeans(dev_sub[, !is_naive, drop = FALSE])
  res$zdiff <- rowMeans(z_sub[, is_naive, drop = FALSE]) -
    rowMeans(z_sub[, !is_naive, drop = FALSE])
  res[order(res$p_value_adjusted), ]
}

tc_res <- compare_subsets(immuneDeviations, "Tc-Naive", "Tc-Mem")
th_res <- compare_subsets(immuneDeviations, "Th-Naive", "Th-Mem")

head(tc_res, 8)
```

## Choosing an effect-size threshold

The helper above returns two effect sizes on different scales. `diff` is
the difference in raw deviation scores; `zdiff` is the same contrast
expressed in Z-score units. A threshold must be applied on the scale it
was defined for, since the two ranges differ by an order of magnitude.

```{r scales}
range(tc_res$diff)
range(tc_res$zdiff)

sum(tc_res$p_value_adjusted < 0.05)
sum(tc_res$p_value_adjusted < 0.05 & abs(tc_res$zdiff) > 0.5)
```

## Visualising the differential motifs

The motifs with the strongest change in the CD8 comparison are shown as
row-wise Z-scores across all four subsets. Samples are ordered by
subset, and the annotation row uses the palette defined above. The fill
scale is diverging and encodes the Z-score, which is a property of the
motif rather than of the subset.

```{r heatmap, fig.height = 6.8, fig.width = 7}
top_motifs <- head(
  tc_res$motifs[
    tc_res$p_value_adjusted < 0.05 & abs(tc_res$zdiff) > 0.5
  ], 25
)

sample_order <- order(colData(immuneDeviations)$cell_type)
z_top <- z_mat[top_motifs, sample_order, drop = FALSE]
sample_levels <- colnames(z_top)
subset_of <- colData(immuneDeviations)$cell_type[sample_order]

heat_df <- data.frame(
  motif = factor(
    rep(rownames(z_top), times = ncol(z_top)),
    levels = rev(top_motifs)
  ),
  sample = factor(
    rep(sample_levels, each = nrow(z_top)),
    levels = sample_levels
  ),
  z = as.vector(z_top)
)

ann_df <- data.frame(
  sample = factor(sample_levels, levels = sample_levels),
  cell_type = subset_of
)

ggplot() +
  geom_tile(data = heat_df, aes(sample, motif, fill = z)) +
  geom_point(
    data = ann_df,
    aes(sample, y = length(top_motifs) + 1.2, colour = cell_type),
    shape = 15, size = 2.4
  ) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, name = "Z-score"
  ) +
  scale_colour_manual(values = cell_type_colors, name = NULL) +
  labs(x = NULL, y = NULL) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 7),
    legend.position = "right"
  )
```

Most of the selected motifs belong to the AP-1 family. Their pattern is
consistent across both T compartments, high in the naive subsets and low
in the memory subsets, and lowest of all in monocytes. The T-box motifs
`EOMES`, `TBR1` and `TBX2` follow a different pattern, low in cytotoxic
memory T cells and in NK cells, which is where those factors are active.
Including the non-T lineages makes clear that the memory signature is
not simply a general lymphoid-myeloid contrast.

# 3. Comparing the two compartments

The CD8 and CD4 contrasts are computed from disjoint sets of samples.
Plotting the Z-score differences against each other shows whether the
two compartments identify the same motifs.

```{r agreement}
shared <- intersect(tc_res$motifs, th_res$motifs)
agree <- data.frame(
  motifs = shared,
  tc = tc_res$zdiff[match(shared, tc_res$motifs)],
  th = th_res$zdiff[match(shared, th_res$motifs)]
)
agree$sig_both <-
  tc_res$p_value_adjusted[match(shared, tc_res$motifs)] < 0.05 &
    th_res$p_value_adjusted[match(shared, th_res$motifs)] < 0.05

rho <- cor(agree$tc, agree$th)
round(rho, 3)
sum(agree$sig_both)
```

```{r agreeplot, fig.height = 4.6}
# Label the extremes at both ends: the strongest motifs cluster tightly,
# so ranking by magnitude alone places every label in one corner.
ranked <- agree[order(agree$tc + agree$th), ]
labelled <- rbind(head(ranked, 4), tail(ranked, 4))

ggplot(agree, aes(tc, th)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(colour = sig_both), size = 1.8, alpha = 0.85) +
  geom_text(
    data = labelled, aes(label = motifs),
    size = 3, vjust = -0.8, check_overlap = TRUE
  ) +
  scale_colour_manual(
    values = c("FALSE" = "grey75", "TRUE" = "#4292C6"),
    name = "significant in both"
  ) +
  labs(
    x = "CD8: Z-score difference (naive - memory)",
    y = "CD4: Z-score difference (naive - memory)",
    title = sprintf("Pearson r = %.2f", rho)
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title.x = element_text(colour = cell_type_colors[["Tc-Mem"]]),
    axis.title.y = element_text(colour = cell_type_colors[["Th-Mem"]])
  )
```

The effect sizes are correlated and share sign for most motifs, so the
same factors are recovered in both compartments.

# 4. Motif footprints

A deviation score summarises a footprint: methylation at the motif
centre relative to its flanking regions, corrected for GC content.
Plotting the footprint shows the profile the score is derived from.

The footprint functions require base-resolution methylation calls and
the motif annotation. This section therefore uses the small `BATF`
example data bundled with the package rather than the pseudobulk object
above. `BATF` is an AP-1 family factor and appears among the motifs
identified in the comparisons above.

The bundled methylome is a single memory T helper sample, so the
observed curve is drawn in that subset's colour from the palette used
throughout this vignette, while the expected curve is kept neutral.

```{r fpdata}
load(system.file("extdata", "BATF_tf_bindsites.rda", package = "methylTFR"))
load(system.file("extdata", "BATF_gcfreqs.rda", package = "methylTFR"))
load(system.file("extdata", "gcdist_BATF.rda", package = "methylTFR"))
load(system.file("extdata", "msites_sub.rda", package = "methylTFR"))

# The bundled methylome is a single sample of one subset, so it is
# labelled and coloured with the same palette used above.
batf_cell_type <- "Th-Mem"
batf_colour <- cell_type_colors[[batf_cell_type]]
```

## Observed and expected profiles

`plotExpectedFootprint()` draws two curves: the methylation observed
around the motif, and the level expected from the GC content of the same
windows. The difference between them is the quantity the deviation score
captures.

```{r expfootprint, fig.height = 4.2, message = FALSE}
plotExpectedFootprint(
  motif = "BATF",
  tf_bindsites = tf_bindsites,
  msites = msites_sub,
  sample_name = batf_cell_type,
  gc_dist = gcdist,
  gcfreqs = gcfreqs,
  enhancer = NULL,
  returnPlotData = FALSE
) +
  scale_colour_manual(values = c(
    Observed = batf_colour, Expected = "grey55"
  ))
```

## Bias-corrected footprint

`plotMotifFootprint()` combines the two profiles into a single corrected
curve, normalised against the outer flanking windows. With
`method = "substraction"` the curve is the observed minus the expected
profile; `method = "division"` uses their ratio.

```{r fpdiff, fig.height = 4.2, message = FALSE}
plotMotifFootprint(
  motif = "BATF",
  tf_bindsites = tf_bindsites,
  msites = msites_sub,
  sample_name = batf_cell_type,
  gc_dist = gcdist,
  gcfreqs = gcfreqs,
  enhancer = NULL,
  method = "substraction",
  flankNorm = 50
) +
  scale_colour_manual(values = batf_colour)
```

A depression at the motif centre indicates methylation below the level
predicted by GC content, which corresponds to a negative deviation
score.

# References

Gündüz IB, Wei B, Chen DC, Wang W, Hariharan M, Norell T, Broderick TJ,
McClain MT, Satterwhite LL, Burke TW, Petzold EA, Shen X, Woods CW,
Fowler VG Jr, Ruffin F, Panuwet P, Barr DB, Wilk AJ, Lee MJ, Blish C,
Castellino F, Walley AM, Evans T, Ecker JR, Müller F, Greenleaf WJ.
Dissecting epigenome dynamics in human immune cells upon viral and
chemical exposure by multimodal single-cell profiling. *bioRxiv*
2025.09.09.675101. doi:
[10.1101/2025.09.09.675101](https://doi.org/10.1101/2025.09.09.675101)

# Session Information

```{r session-info}
sessionInfo()
```
