---
title: "methylTFR: Quantification of DNA Methylation Patterns in TFBS"
author: 
  - name: Irem B. Gunduz
    affiliation:
    - Saarland University
    email: irembgunduz@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('methylTFR')`"
vignette: >
  %\VignetteIndexEntry{methylTFR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
  R = citation(),
  BiocStyle = citation("BiocStyle")[1],
  knitr = citation("knitr")[1],
  RefManageR = citation("RefManageR")[1],
  rmarkdown = citation("rmarkdown")[1],
  sessioninfo = citation("sessioninfo")[1],
  testthat = citation("testthat")[1],
  methylTFR = citation("methylTFR")[1]
)
```

# Basics

## Install `methylTFR`

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. `r Biocpkg("methylTFR")` is a `R` package available via the [Bioconductor](http://bioconductor.org) repository for packages. `R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install `r Biocpkg("methylTFR")` by using the following commands in your `R` session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("methylTFR")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

# Introduction to `methylTFR`
#TODO add more details

## Annotation data from `methylTFRAnnotation` packages
We provided annotation package for **Hg38**, `r Biocpkg("methylTFRAnnotationHg38")`, which contains the annotation files for `r Biocpkg("methylTFR")` to run.
If you want create your own `r Biocpkg("methylTFR")` annotation packages for other genomes, we recommend visiting [methylTFRAnnotationCreator](https://github.com/EpigenomeInformatics/methylTFRAnnotationCreator) project.

Breifly, `run_methyltfr` takes following arguments from annotation package:
- `tf_bindsites`: TF binding sites in GRanges format
- `gcfreqs`: GC frequency table list
- `gc_dist`: GC distribution *GRanges* object for each chromosome

Here is a subsetted example of annotation data from `r Biocpkg("methylTFRAnnotationHg38")` package:

```{r "annotation", message=FALSE}
load(system.file("extdata", "gcdist_subset.rda", package = "methylTFR"))
load(system.file("extdata", "FOXF2_gcfreqs.rda", package = "methylTFR"))
load(system.file("extdata", "FOXF2_tf_bindsites.rda", package = "methylTFR"))

# Subsetted GC distribution data for chr1
head(gcdist)

# Subsetted GC frequency data for FOXF2 motif
head(gcfreqs$FOXF2[, 1:5])

# Subsetted FOXF2 motif binding sites
head(tf_bindsites$FOXF2)
```

## Importing data using `methylTFR`

`r Biocpkg("methylTFR")` provides `read_methylome` function to import data from various sources.
Currently, `read_methylome` supports importing data from *EPP*, *bissnp*, *bismarkCytosine*, *bismarkcov*, *allc* and *encode* file formats.

Here is a quick example of how to import data from *EPP* file format.

```{r "read_methylome", message=FALSE, eval = FALSE}
# Import data from EPP file format
msites <- read_methylome(
  filename = "sample.bed", # EPP file path
  filetype = "EPP" # file type
)
```
`read_methylome` then returns a *GRanges* object with following metadata columns:
- `score`: methylation score 
- `coverage`: coverage

```{r "import", message=FALSE}
# Load sample data
load(system.file("extdata", "example_data.rda", package = "methylTFR"))

# Subsetted sample data
head(msites)
```

## Computing deviation scores with `methylTFR`

Here is a quick example of how to compute bias-corrected deviation scores with `r Biocpkg("methylTFR")` for motif *FOXF2* from **JASPAR2020* motifset in subsetted sample from [EGAS00001001624](https://ega-archive.org/studies/EGAS00001001624).

```{r "computedevs", message=FALSE}
library(methylTFR)

# Compute deviation scores
devs <- computeDeviation("FOXF2", # motif name
  msites = msites, # sample data
  tf_bindsites = tf_bindsites, # TF binding sites
  gcfreqs = gcfreqs, # GC frequency
  gcdist = gcdist, # GC distribution
  ignoreStrand = TRUE
)

# Bias-corrected deviation score for FOXF2
devs
```

Alternatively, if you want to access expected and bias corrected deviation scores for a motif, you can use `computeDeviation` with `intermediate = TRUE` argument.

```{r "computedevs2", message=FALSE}
# Compute deviation scores
devs <- computeDeviation("FOXF2", # motif name
  msites = msites, # sample data
  tf_bindsites = tf_bindsites, # TF binding sites
  gcfreqs = gcfreqs, # GC frequency
  gcdist = gcdist, # GC distribution
  ignoreStrand = TRUE, # ignore strand information
  intermediate = TRUE # return intermediate results
)

# Expected and bias corrected deviation score for FOXF2
devs
```

# Quick start to using `methylTFR` for multiple samples

We implemented `run_methyltfr` function, to run `r Biocpkg("methylTFR")` on sets of samples. `run_methyltfr` takes following arguments:

```{r "run_methyltfr", message=FALSE, eval = FALSE}
library(GenomicRanges)
library(dplyr)
library(methylTFRAnnotationHg38) # annotation package for hg38
library(methylTFR)

gcfreqs <- getGCfreq(motifSet = "jaspar2020")
gc_dist <- getGenomeGC()
tf_bindsites <- getTFbindsites(motifSet = "jaspar2020")

sample_dir <- file.path("samples_dir")
sample_ann <- "samples.tsv" # should contain column name bedFile

# deviation score matrix
deviations <- run_methyltfr(sample_ann, # sample annotation file
  sample_dir, # where the EPP files are
  threads = 8, # number of threads
  chunkSize = 10, # number of chunks to process
  sampleColName = "bedFile", # column name for EPP file paths in sample_ann
  tf_bindsites = tf_bindsites, # TF binding sites
  gcfreqs = gcfreqs, # GC frequency
  gc_dist = gc_dist, # GC distribution
  filetype = "EPP" # file type
)
```

# Citing `methylTFR`

We hope that `r Biocpkg("methylTFR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("methylTFR")
```

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `methylTFR` tag and check [the older posts](https://support.bioconductor.org/tag/methylTFR/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

## Reproducibility

The `r Biocpkg("methylTFR")` package `r Citep(bib[["methylTFR"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("methylTFR.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("methylTFR.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

## Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
